New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-12 jso.qmd
use the ferrn pkg to make the huber plot

```{r}
#| label: huber-plot-function
#| echo: false
huber_plot <- function (origdata2D, origclass=NULL, 
  PPmethod = "LDA", weight = TRUE, r = 1, lambda = 0.5, 
  opt.proj = TRUE, UserDefFtn = NULL, ...) 
{
    index <- NULL
    best.proj <- NULL
    best.index <- 0
    origdata2D <- as.matrix(origdata2D)
    for (i in 0:360) {
        theta <- pi/180 * i
        proj.data <- matrix(cos(theta) * origdata2D[, 1] + sin(theta) * 
            origdata2D[, 2])
        proj <- matrix(c(cos(theta), sin(theta)), ncol = 1)
        if (PPmethod == "LDA") {
            newindex <- PPtreeViz::LDAindex(origclass, origdata2D, proj = proj, 
                weight = weight)
        else if (PPmethod == "PDA") {
            newindex <- PPtreeViz::PDAindex(origclass, origdata2D, proj, 
                weight = weight, lambda = lambda)
        else if (PPmethod == "Lr") {
            newindex <- PPtreeViz::Lrindex(origclass, origdata2D, proj, 
                weight = weight, r = r)
        else if (PPmethod == "GINI") {
            newindex <- PPtreeViz::GINIindex1D(origclass, origdata2D, proj)
        else if (PPmethod == "ENTROPY") {
            newindex <- PPtreeViz::ENTROPYindex1D(origclass, origdata2D, 
                proj)
        else if (PPmethod == "holes") {
          newindex <- tourr::holes()(proj.data)
        else if (PPmethod == "skewness") {
          newindex <- tourr::skewness()(proj.data)
        else if (PPmethod == "norm_bin") {
          newindex <- tourr::norm_bin(nrow(proj.data))(proj.data)
        else if (PPmethod == "UserDef") {
            newindex <- UserDefFtn(proj.data, ...)
        index <- c(index, newindex)
    sel.index <- which(index[1:360] > signif(max(index), 6) - 
        1e-06)
    theta.best.all <- pi/180 * (sel.index - 1)
    theta.best <- theta.best.all[1]
    proj.data.best <- matrix(cos(theta.best) * origdata2D[, 1] + 
        sin(theta.best) * origdata2D[, 2])
    index.best <- max(index)
    range <- round(max(index) - min(index), 5)
    if (range == 0) {
      PPindex <- rep(4, length(index))
    else {
      PPindex <- (index - min(index))/range * 2 + 3
    data.circle <- NULL
    data.index <- NULL
    for (i in 1:361) {
        theta <- pi/180 * (i - 1)
        data.index <- rbind(data.index, c(PPindex[i] * cos(theta), 
            PPindex[i] * sin(theta)))
        data.circle <- rbind(data.circle, c(4 * cos(theta), 4 * 
            sin(theta)))
    maxdiff <- max(c(diff(range(origdata2D[, 1])), diff(range(origdata2D[, 
        2]))))
    orig.scaled <- apply(origdata2D, 2, function(x) (x - mean(x))/maxdiff * 
        3.5)
    data.cX <- data.circle[, 1]
    data.cY <- data.circle[, 2]
    data.X <- data.index[, 1]
    data.Y <- data.index[, 2]
    plot.data <- data.frame(data.cX, data.cY, data.X, data.Y)
    x <- orig.scaled[, 1]
    y <- orig.scaled[, 2]
    if (is.null(origclass)) 
      group <- rep(1, nrow(origdata2D))
    else
      group <- origclass
    point.data <- data.frame(x, y, group)
    min.X <- min(unlist(plot.data))
    max.X <- max(unlist(plot.data))
    if (is.null(origclass)) {
      P1 <- ggplot(data = plot.data, aes(x = data.X, y = data.Y)) + 
        geom_path() + geom_path(aes(x = data.cX, y = data.cY), 
        linetype = "dashed") + geom_point(data = point.data, 
        aes(x = x, y = y)) + scale_x_continuous(breaks = NULL) + 
        scale_y_continuous(breaks = NULL) + xlab("") + ylab("") + 
        coord_fixed() + theme_bw() + theme(panel.border = element_blank())
    } else {
      P1 <- ggplot(data = plot.data, aes(x = data.X, y = data.Y)) + 
        geom_path() + geom_path(aes(x = data.cX, y = data.cY), 
        linetype = "dashed") + geom_point(data = point.data, 
        aes(x = x, y = y, color = group, shape = group)) +
        scale_x_continuous(breaks = NULL) + 
        scale_y_continuous(breaks = NULL) + xlab("") + ylab("") + 
        coord_fixed() + theme_bw() + theme(panel.border = element_blank())
    if (opt.proj) {
        P1 <- P1 + geom_abline(intercept = 0, slope = sin(theta.best)/cos(theta.best))
        if (length(theta.best.all) > 1) 
            for (i in 2:length(theta.best.all)) P1 <- P1 + geom_abline(intercept = 0, 
                slope = sin(theta.best.all[i])/cos(theta.best.all[i]), 
                linetype = "dashed")
    best.proj.data <- proj.data.best
    #group <- origclass
    if (is.null(origclass)) {
      hist.data <- data.frame(best.proj.data)
      P2 <- ggplot(data = hist.data, aes(x = best.proj.data)) + 
        geom_histogram()
    } else {
      hist.data <- data.frame(best.proj.data, group)
      P2 <- ggplot(data = hist.data, aes(x = best.proj.data, 
                                         group = group)) + 
        geom_histogram(aes(fill = group), position = "stack")
    #gridExtra::grid.arrange(P1, P2, nrow = 1)
    return(list(plot=P1, proj_data=best.proj.data))

@@

```{r}
#| label: fig-example-functions
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 80%
#| fig-cap: "Examples of PP indexes with large (top row) and small (bottom row) squint angles, shown with a Huber plot, and histogram of the projected data corresponding to the optimal projection. A Huber plot shows the PP index values for all 1D data projections in polar coordinates."
data(randu)
randu_std <- as.data.frame(apply(randu, 2, function(x) (x-mean(x))/sd(x)))
randu_std$yz <- sqrt(35)/6*randu_std$y-randu_std$z/6

#proj <- animate_xy(flea[,1:6], guided_tour(lda_pp(flea$species)))
best_proj <- matrix(c(-0.44157005,  0.80959831,
                      0.14622089, -0.08228193,
                      -0.08454314, -0.23234287,
                      0.70367852,  0.28185021,
                      0.27876825,  0.44819691,
                      0.45123453,  0.05896646), ncol=2, byrow=T)
flea2D <- as.matrix(flea[,1:6]) %*% best_proj
colnames(flea2D) <- c("X1", "X2")
flea2D <- as.data.frame(flea2D)

# Make both plots
large_sqa <- huber_plot(flea2D, PPmethod="skewness")
small_sqa <- huber_plot(randu_std[c(1,4)], PPmethod="norm_bin")

sqa1 <- large_sqa$plot + ggtitle("(a) Skewness index")
lsqa_d <- tibble(d = large_sqa$proj_data[,1])
sqa2 <- ggplot(lsqa_d, aes(x=d)) + 
  geom_histogram(binwidth=0.25, colour="white") +
  xlab("") + ylab("") +
  ggtitle("(b) Optimally projected data") +
  theme_bw() +
  theme(axis.text.y = element_blank())
sqa3 <- small_sqa$plot + ggtitle("(c) Binned normality index")
ssqa_d <- tibble(d = small_sqa$proj_data[,1])
sqa4 = ggplot(ssqa_d, aes(x=d)) + 
  geom_histogram(breaks = seq(-2.2, 2.4, 0.12)) +
  xlab("") + ylab("") +
  ggtitle("(d) Optimally projected data") +
  theme_bw() +
  theme(axis.text.y = element_blank())

sqa1 + sqa2 + sqa3 + sqa4 + plot_layout(ncol=2, widths=c(2,2))
```

@@

Here, the potential for a Jellyfish Search Optimiser (JSO) (see @chou_novel_2021, and @rajwar_exhaustive_2023) for the PPGT is explored. JSO, inspired by the search behaviour of jellyfish in the ocean, is a swarm-based metaheuristic designed to solve global optimisation problems. Compared to traditional methods, JSO has demonstrated stronger search ability and faster
convergence, and requires fewer tuning parameters. These practicalities
make JSO a promising candidate for enhancing PP optimisation.
<!--The study reported here explores JSO's potential in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.-->

@@

<!-- Need to assume that the special issue will have the general introduction to PP, so the focus here will be on background needed for this paper, tours, guidance using PP, optimisation, visualisation, ... -->

>>>>

```{r}
#| label: fig-example-functions
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 80%
#| fig-cap: "Examples of PP indexes with large (top row) and small (bottom row) squint angles, shown with a Huber plot, and histogram of the projected data corresponding to the optimal projection. A Huber plot shows the PP index values for all 1D data projections in polar coordinates."
#proj <- animate_xy(flea[,1:6], guided_tour(lda_pp(flea$species)))
best_proj <- matrix(c(-0.44157005,  0.80959831,
                      0.14622089, -0.08228193,
                      -0.08454314, -0.23234287,
                      0.70367852,  0.28185021,
                      0.27876825,  0.44819691,
                      0.45123453,  0.05896646), ncol=2, byrow=T)

flea2D <- as.matrix(flea[,1:6]) %*% best_proj
colnames(flea2D) <- c("X1", "X2")
flea2D <- as.data.frame(flea2D)
flea_huber <- prep_huber(flea2D, index = skewness())
sqa1 <- ggplot() +
  geom_huber(data = flea_huber$idx_df, aes(x = x, y = y)) +
  geom_point(data = flea2D, aes(x = X1, y = X2)) +
  geom_abline(slope = flea_huber$slope, intercept = 0) + 
  coord_fixed() +
  theme_huber() +
  ggtitle("(a) Skewness index")

sqa2 <- ggplot(flea_huber$proj_df, aes(x=x)) + 
  geom_histogram(binwidth=0.25, colour="white") +
  xlab("") + ylab("") +
  ggtitle("(b) Optimally projected data") +
  theme_bw() +
  theme(axis.text.y = element_blank())

data(randu)
randu_std <- as.data.frame(apply(randu, 2, function(x) (x-mean(x))/sd(x)))
randu_std$yz <- sqrt(35)/6*randu_std$y-randu_std$z/6
randu_df <- randu_std[c(1,4)]
randu_huber <- prep_huber(randu_df, index = norm_bin(nr = nrow(randu_df)))

sqa3 <- ggplot() +
  geom_huber(data = randu_huber$idx_df, aes(x = x, y = y)) +
  geom_point(data = randu_df, aes(x = x, y = yz)) +
  geom_abline(slope = randu_huber$slope, intercept = 0) +
  theme_huber() +
  coord_fixed() + 
  ggtitle("(c) Binned normality index")

sqa4 <- ggplot(randu_huber$proj_df, aes(x = x)) +
  geom_histogram(breaks = seq(-2.2, 2.4, 0.12)) +
  xlab("") + ylab("") +
  theme_bw() +
  theme(axis.text.y = element_blank()) + 
  ggtitle("(d) Optimally projected data")

sqa1 + sqa2 + sqa3 + sqa4 + plot_layout(ncol=2, widths=c(2,2))

@@

@@

Here, the potential for a Jellyfish Search Optimiser (JSO) (see @chou_novel_2021, and @rajwar_exhaustive_2023) for the PPGT is explored. JSO, inspired by the search behaviour of jellyfish in the ocean, is a swarm-based metaheuristic designed to solve global optimisation problems. Compared to traditional methods, JSO has demonstrated stronger search ability and faster
convergence, and requires fewer tuning parameters. These practicalities
make JSO a promising candidate for enhancing PP optimisation.

@@

### 2024-09-12 jso.tex
use the ferrn pkg to make the huber plot

\begin{tcolorbox}[enhanced jigsaw, colback=white, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, left=2mm, arc=.35mm, breakable, bottomrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, opacitybacktitle=0.6, toprule=.15mm, titlerule=0mm, leftrule=.75mm]

>>>>

\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colback=white, opacityback=0, left=2mm, toprule=.15mm, breakable, leftrule=.75mm, toptitle=1mm, bottomtitle=1mm, titlerule=0mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
